### 2018-08-14 doc/intro.tex
gsa abstract

>>>>

One of the great promises of paleobiology is that by studying the past we can better predict the future.  This promise is particularly pertinent given suggestions that risk assessments for some modern species could be improved by examining past extinction patterns and by using paleontological records to establish geographic range and abundance trajectories on geological timescales.  Any effort to assess future risk based on past extinctions and range trajectories must address two key questions:  (1) At a given timescale, are geographic range trajectories deterministic (past trends are likely to continue into the future) or Markovian (the future depends only on the present state)? (2) Given knowledge of past extinction/survival patterns and the present geographic ranges of extant taxa, how accurate are extinction risk predictions?

To address these questions we analyze the fossil record of Cenozoic planktonic microfossil taxa (foramanifera, radiolarians, diatoms, and calcareous nanoplankton). Using a model of species survival, we analyzed how survival probability changes over time as a function of species age, time of observation, current geographic range, most recent change in geographic range, global temperature average, and the lag of global temperature. Our best supported model includes the historical covariates, change in geographic range and lag of global temperature, which indicates that the past improves our estimates of the present and future.

The effects of the historical covariates are extremely small and vary considerably over time. For example, the effect of change in geographic range size can either increase or decrease probability of extinction depending on when in the Cenozoic the observation takes place. The improvement in predictive power by including these historical covariates is modest at best and reflects the rarity of extinction events (i.e. class-imbalance) and the extremely stochastic nature of species survival. Correcting for class imbalance, we find that in-sample model performance measures are approximately equal to out-of-sample performance as estimated from cross-validation. These results reflect the difficulty of estimating species extinction, and that while including historical covariates does improve model performance, that gain is very small.  This result implies that at million-year timescales geographic range trajectories are nearly Markovian, perhaps because the processes driving geographic range changes vary on substantially shorter timescales. The effect of change in geographic range on survival most likely stands for many interacting and unobserved processes which in-turn produce that species' geographic range and its affect on survival.

Finally, we find support for species' extinction risk increases with age, though the strength of this effect varies among taxonomic groups. This effect is most pronounced in forams and radiolarians, and less pronounced in diatoms and calcareous nanoplankton. The greatest source of variance in survival probability is the timing of observation. Importantly, this result means that the time of an observation is a greater source of variation in survival probability when compared to species age.

Ultimately, we find that including information on a species' change in geographic range size on average improves our predictions of species survival at million-year timescales. However, the effect of change in geographic range is much smaller than the effect of current geographic range, and highly variable through time as the effect changes sign and there are times where there is little evidence for any effect of past geographic range. The results of this study reinforce the importance of the promise of paleontology and using the past to predict the future.

how predictable is extinction? how much does the past improve our predictions? how much does the past improve our predictions?

we know that at almost all times, a species with geater than average geographic range is expected to have a lower extinction risk than a closely related species with a lower than average geographic range. what we don't know is if expanding or decreasing in geographic range over million-year time scales has any additional predictive information about a species' risk of going extinct.

### 2018-08-14 doc/methods.tex
no interaction model
gsa abstract

Data was binned into 1 My bins based on the estimated age of the fossil occurrence. The estimated ages of each occurrence is a product of core-specific age-models and are overly precise. The hope is that binning the data overcomes the heterogenity in age-models between the cores. 
% binning, [t - 1, t)

@@

% lots of ways of varying
% maxgcd * diff_maxgcd varies first by when then by fossil group.
% non-nested varying intercept based on age, then by fossil group.

% event ~ temp + lag1_temp + maxgcd * diff_maxgcd + 
% (1 + maxgcd * diff_maxgcd | fact_mybin/fossil.group) +
% (1 | fact_relage/fossil.group)
The logistic regression can thus be expressed as 
% scripts
%   t = death? 0 if not last observation
%   \beta = effect range vars (varies over time and by fossil group)
%   \gamma = effect temp vars
%   a = age effect
% subscripts
%   w = when in My
%   f = fossil group
%   d = age at observation
% 
\begin{equation}
  \begin{aligned}
    t_{i} &= logit^{-1}(\beta_{w[i], f[i]} + \beta_{w[i], f[i]} range_{i} + \beta_{w[i], f[i]} rangediff_{i} + \beta_{w[i], f[i]} range_{i} rangediff_{i} + \gamma_{1} temp_{w[i]} + \gamma_{2} templag_{w[i]} + a_{d[i], f[i]}) \\
    \beta_{w, f} &\sim MVN(\alpha_{f}, \Sigma) \\
    \alpha_{f} &\sim MVN(0, \Sigma) \\
    \gamma &\sim N(0, \sigma) \\
    a_{w, f} &\sim MVN(\delta_{f}, \Sigma) \\
    \delta_{f} &\sim MVN(0, \Sigma) \\
  \end{aligned}
  \label{eq:core}
\end{equation}
with \(i\) indexing the observation and bracket subscripts referencing the class of the \(i\)th observation e.g. \(j[i]\) is the age of the \(i\)th observation. \(a_{j[i]}\) is taxon age in bins at time of observation where \(j = 1, 2, \cdots, J\) and \(J\) is the maximum observed age of any taxon. \(b_{k}\) is the time of observation where \(k = 1, 2, \cdots, K\) and \(K\) is the number of time bins observed. Finally, \(c_{f}\) is the taxonomic group of an observation where \(f = 1, 2, \cdots, F)\) and \(F\) is the total number of taxonomic groups analyzed.

@@

Posterior convergence was determined using the following general and HMC specific diagnostic criteria: scale reduction factor (\(\hat{R}\); target \(<1.1\)), effective sample size (eff; target value eff/steps \(<0.0001\)), number of samples that saturated the maximum trajectory length for avoiding infinite loops (treedepth; target value 0), sample divergence, and the energy Bayesian Fraction of Mission Information (E-BFMI; target value \(>0.2\)). For futher explanation of these diagnostic criteria, see the Stan Manual cITATION.

@@

We considered four variants of this model: 1) only age-based varying intercepts, 2) age and time of observation varying intercepts, 3) age and taxonomic group varying intercepts, and 4) all three considered varying intercepts. These models were compared using three measures of model performance: WAIC and LOOIC which are estimates of a model's out-of-sample performance, and ROC which is a measure of a model's predictive performance.

@@

>>>>

Data was binned into 1 My bins based on the estimated age of the fossil occurrence. The estimated ages of each occurrence is a product of core-specific age-models and are overly precise. The hope is that binning the data overcomes the heterogenity in age-models between the cores.

@@

The logistic regression can thus be expressed as 
\begin{equation}
  \begin{aligned}
    t_{i} &= logit^{-1}(\beta_{w[i], f[i]} + \beta_{w[i], f[i]} range_{i} + \beta_{w[i], f[i]} rangediff_{i} + \gamma_{1} temp_{w[i]} + \gamma_{2} templag_{w[i]} + a_{d[i], f[i]}) \\
    \beta_{w, f} &\sim MVN(\alpha_{f}, \Sigma) \\
    \alpha_{f} &\sim MVN(0, \Sigma) \\
    \gamma &\sim N(0, \sigma) \\
    a_{w, f} &\sim MVN(\delta_{f}, \Sigma) \\
    \delta_{f} &\sim MVN(0, \Sigma) \\
  \end{aligned}
  \label{eq:core}
\end{equation}
with \(i\) indexing the observation and bracket subscripts referencing the class of the \(i\)th observation e.g. \(j[i]\) is the age of the \(i\)th observation. \(a_{j[i]}\) is taxon age in bins at time of observation where \(j = 1, 2, \cdots, J\) and \(J\) is the maximum observed age of any taxon. \(b_{k}\) is the time of observation where \(k = 1, 2, \cdots, K\) and \(K\) is the number of time bins observed. Finally, \(c_{f}\) is the taxonomic group of an observation where \(f = 1, 2, \cdots, F)\) and \(F\) is the total number of taxonomic groups analyzed.

@@

Posterior convergence was determined using the following general and HMC specific diagnostic criteria: scale reduction factor (\(\hat{R}\); target \(<1.1\)), effective sample size (eff; target value eff/steps \(<0.0001\)), number of samples that saturated the maximum trajectory length for avoiding infinite loops (treedepth; target value 0), sample divergence, and the energy Bayesian Fraction of Mission Information (E-BFMI; target value \(>0.2\)). For futher explanation of these diagnostic criteria, see the Stan Manual CITATION.

@@

We considered four variants of this model: 1) historical covariates with varying intercepts and slopes, 2) no historical coviarates with varying intercepts and slopes, 3) historical covariates without varying intercepts and slopes, and 4)

@@

\subsection{Prediction given class imbalance}

When the classes are not approximately balanced, the base probability of 1 versus 0 is below 0.5. When the intercept term of the logistic regression is sufficiently negative (-2 or less), the only way to obtain a linear predictor value of 0.5 or greater is if the covariate effects are large or there the covariates are at extreme values. The 0.5 default cut-off point for binary prediction is much to high. A method for determining the new ``optimal'' cutpoint involves calculating the ROC curve for the data and finding the point along that curve which maximizes both true positive rate and the false positive rate; the combination with the greatest sum corresponds to the new optimal cutpoint CITATION. Class membership is then predicted given this new cutpoint. This process is done for every posterior draw, as each realization from the posterior predictive distribution has its own optimal cutpoint.

\subsection{Cross-validation}

Cross-validation is a procedure for estimating the expected out-of-sample predictive error of a model CITATION. By dividing the data into sub-components (folds), fitting a model to some of the folds folds, and then predicting the response values for the held-out fold. For our dataset, expected out-of-sample AUC was estimated using five-fold cross-validation. For time-series data, the folds are approximately equal segments of time. The model is fit to the first fold and the posterior estimates are used to predict the states of the observations in the second fold, then the model is fit to the first and second fold and the posterior states are used to estimate the states from the third fold, and so on with increasingly large numbers of folds used for fitting a model to predict the states from the subsequent fold. With 63 time points, each of the five folds represents approximately 13 time points. Keep in mind, however, that each time point is represented by many (100-1000) individual observations.

For each posterior draw of the cross-validation predictions, the optimal cutpoint was calculated as described above. Given these class assignments based on estimated optimal cutpoint, the AUC value of the out-of-sample estmates was calculated. AUC values are calculated for the entire fold and for each of the individual time points represented within that fold.

### 2018-08-14 doc/results.tex
gsa abstract

\section{Data Specifications}

Microfossil occurrence data was downloaded from the Neptune Database \url{http://www.nsb-mfn-berlin.de/nannotax}. Occurrence information was downloaded for calcareous nannofossils, diatoms, dinoflaggellants, foraminifera, and radiolarians from the entire globe and having occurred between 120 and 0 Mya. This selection of species was then culled to just those species which have their first occurrence at most 63 My; this avoids those taxa which survived the K/Pg boundary, those taxa arrising just after the K/Pg boundary, and lines up the occurrences with the temperature time-series (discussed below).

Data was binned into 1 My bins based on the estimated age of the fossil occurrence. The estimated ages of each occurrence is a product of core-specific age-models and are overly precise. The hope is that binning the data overcomes the heterogenity in age-models between the cores.

The survival or extinction of a taxon is determined for each taxons' occurrences. For every occurrence except the last, the taxon has survived which is indicated by a 0. The last occurrence of the taxon is considered the bin in which the taxon has gone extinct. This protocol means that we are reading the fossil record ``as written,'' a practice that is potentially dangerous CITATIONS but is common with marine microfossil data CITATIONS.

A taxon's geographic range during a 1 My bin was calculated as the maximum great circle distance on an ellipsoid between any two occurrences, also called a geodesic. Geographic range was measured in kilometers.

Geographic range was then log-plus-one transformed and then standardized by zero-centering the data and then dividing by its standard deviation so that geographic range had mean 0 and standard deviation 1.

Temperature data is based on estimates from Cramer et al CITATION. These estimates are based on Magnesian/Calcium isotope ratios and are a more accurate estimate of global temperature than the standard Oxygen isotope-based estimates because Mg/Ca based estimates are not effected by ice-volume and fresh-water input (e.g. metioric water). Cramer et al. provide temperature estimates for every 0.1 My from 0 to 63 Mya; we binned these estimates every 1 My in accordance with the way we binned all fossil occurrences.

\section{Model Specifications}

In survival analysis, the hazard funciton describes the instanenous rate of extinction at time \(t\) given that the species has lived up to that point. For discrete-time survival data, this is expressed as \(\lambda(t | X) = P(T = t | T \geq t, X)\) for \(t = 1, \cdots N\) where \(t\) is the status of a species at a time point, \(T\) is the time of extinction of the species, and \(i = 1, 2, \cdots, N\) with \(N\) being the total number of observations.

In survival analysis the hazard function describes the in stannous rate of extinction for a species given matrix of covariates X, \(\lambda(t | X)\). We parameterized the hazard function as a logistic regression/generalized linear model defined \(\lambda(t | X) = h(\Theta)\) where \(h(.)\) is an inverse logit link function and \(\Theta\) is the probability of a taxon going extinction during interval \(t\).

\(\Theta\) is then parameterized as a hierarchical model with multiple non-nested varying intercepts CITATION. First, we considered two aspects of geographic range as continuous covariates: geographic range \(r\) during interval \(t\), and the difference \(d\) between the geographic range at \(t - 1\) and \(t\). We also considered the interaction between geographic range during intervals \(t\) and \(t - 1\). At the first observation for a taxon, the diff of geographic range is 0.

Second, we considered two aspects of global temperature: mean temperature during interval \(t\), and the lag of mean temperature (i.e. mean temperature during interval \(t - 1\).)

The logistic regression can thus be expressed as 
\begin{equation}
  \begin{aligned}
    t_{i} &= logit^{-1}(\beta_{w[i], f[i]} + \beta_{w[i], f[i]} range_{i} + \beta_{w[i], f[i]} rangediff_{i} + \gamma_{1} temp_{w[i]} + \gamma_{2} templag_{w[i]} + a_{d[i], f[i]}) \\
    \beta_{w, f} &\sim MVN(\alpha_{f}, \Sigma) \\
    \alpha_{f} &\sim MVN(0, \Sigma) \\
    \gamma &\sim N(0, \sigma) \\
    a_{w, f} &\sim MVN(\delta_{f}, \Sigma) \\
    \delta_{f} &\sim MVN(0, \Sigma) \\
  \end{aligned}
  \label{eq:core}
\end{equation}
with \(i\) indexing the observation and bracket subscripts referencing the class of the \(i\)th observation e.g. \(j[i]\) is the age of the \(i\)th observation. \(a_{j[i]}\) is taxon age in bins at time of observation where \(j = 1, 2, \cdots, J\) and \(J\) is the maximum observed age of any taxon. \(b_{k}\) is the time of observation where \(k = 1, 2, \cdots, K\) and \(K\) is the number of time bins observed. Finally, \(c_{f}\) is the taxonomic group of an observation where \(f = 1, 2, \cdots, F)\) and \(F\) is the total number of taxonomic groups analyzed.

\section{Model Parameter Estimation}

We implemented our model using the \begin{texttt}rstanarm\end{texttt} package for the R programming language CITATION. This package provides an interface to the Stan probabilistic programming language for writing hierarchical/mixed-effects models in native R. Posterior estimates were obtained through Hamiltonian Monte Carlo, using 2000 steps divided equally between warm-up and sampling. In order to prevent divergent transitions the adapt delta value was increased to 0.99; all other HMC/NUTS sampling parameters were kept at the defaults for rstanarm version XX CITATION.

Posterior convergence was determined using the following general and HMC specific diagnostic criteria: scale reduction factor (\(\hat{R}\); target \(<1.1\)), effective sample size (eff; target value eff/steps \(<0.0001\)), number of samples that saturated the maximum trajectory length for avoiding infinite loops (treedepth; target value 0), sample divergence, and the energy Bayesian Fraction of Mission Information (E-BFMI; target value \(>0.2\)). For futher explanation of these diagnostic criteria, see the Stan Manual CITATION.

\section{Model Selection and Adequacy}

We considered four variants of this model: 1) historical covariates with varying intercepts and slopes, 2) no historical coviarates with varying intercepts and slopes, 3) historical covariates without varying intercepts and slopes, and 4)

To determine how many non-nested varying-intercept terms were possible to include without overly biasing out models' parameter estimates, the four variant models described above were compared using both WAIC and LOOIC which are estimates of a model's expected out-of-sample performance; these measures are fully Bayesian, taking into account the entire estimated joint posterior. CITATION. WAIC and LOOIC are interpreted similarly to AIC, with lower values indicating greater expected out-of-sample performance. Additionally, the Bayesian nature of WAIC and LOOIC mean that they are calculated with standard errors allowing for straight forward comparisons. We selected the model with lowest WAIC and LOOIC and all models within one standard error of its WAIC/LOOIC estimate.

Given these selected models, we did a further comparison between models including the change in geographic range from \(t - 1\) to \(t\), and its interaction with geographic range at \(t\), in the model to determine how much our predictions of extinction risk are improved by the inclusion of ``the past.''

The model adequacy was measured using the area under the reciever operating characteristic curve (AUC). This measure is commonly used in classification problems like this one as it has the desireable characteristic of comparing the model's true positive rate with its false positive rate, as opposed to only true positive rate measured by accuracy CITATION. This value ranges between 0.5 and 1, with 0.5 indicating no improvement in performance from random and 1 indicating perfect performance. AUC was calculated for the dataset as a whole, and for each of time bins.

\subsection{Prediction given class imbalance}

When the classes are not approximately balanced, the base probability of 1 versus 0 is below 0.5. When the intercept term of the logistic regression is sufficiently negative (-2 or less), the only way to obtain a linear predictor value of 0.5 or greater is if the covariate effects are large or there the covariates are at extreme values. The 0.5 default cut-off point for binary prediction is much to high. A method for determining the new ``optimal'' cutpoint involves calculating the ROC curve for the data and finding the point along that curve which maximizes both true positive rate and the false positive rate; the combination with the greatest sum corresponds to the new optimal cutpoint CITATION. Class membership is then predicted given this new cutpoint. This process is done for every posterior draw, as each realization from the posterior predictive distribution has its own optimal cutpoint.

\subsection{Cross-validation}

Cross-validation is a procedure for estimating the expected out-of-sample predictive error of a model CITATION. By dividing the data into sub-components (folds), fitting a model to some of the folds folds, and then predicting the response values for the held-out fold. For our dataset, expected out-of-sample AUC was estimated using five-fold cross-validation. For time-series data, the folds are approximately equal segments of time. The model is fit to the first fold and the posterior estimates are used to predict the states of the observations in the second fold, then the model is fit to the first and second fold and the posterior states are used to estimate the states from the third fold, and so on with increasingly large numbers of folds used for fitting a model to predict the states from the subsequent fold. With 63 time points, each of the five folds represents approximately 13 time points. Keep in mind, however, that each time point is represented by many (100-1000) individual observations.

For each posterior draw of the cross-validation predictions, the optimal cutpoint was calculated as described above. Given these class assignments based on estimated optimal cutpoint, the AUC value of the out-of-sample estmates was calculated. AUC values are calculated for the entire fold and for each of the individual time points represented within that fold.

>>>>

\section{Model choice and adequacy}

% loo table

% waic table

% ROC model comparison

% ROC model comparison time series

\section{Cross-validation}

% ROC OOS estimate

% ROC OOS estimate time series

\section{Parameter estimates}

% effects, average

% effects, time series

% variance components

% risk estimate compared to change in geo-range
